\thetitle

### 2020-11-15 writing/challenges.tex
working on rational

The second big challenge is the disparity of gradually
typed languages. While the approaches are well understood in theory,
they come with vastly different type systems, accommodate seemingly
incompatible language features, and apply to radically different coding
styles in the untyped world.  To eliminate the effect of these differences 
on its results, the application of an evaluation method to different
gradual typing systems must use the same test programs and the same type
system. Otherwise, results for different gradual typing systems 
do not share a common baseline and, therefore, do not lend support to any
conclusions about whether the value of blame changes from one system to
another. For instance, the case study presented here builds on
Greenman and Felleisen's transient variant for functional Typed
Racket~\cite{gf-icfp-2018} to obtain a transient semantics that is
comparable with Typed Racket's default natural one. This transient
variant accommodates many more linguistic features and library calls
in a way that is faithful to the transient
semantics~\cite{vss-popl-2017}.

@@

Supporting these strategies is {\em not\/} just an issue of language
implementation but affects greatly the behavior of the rational programmer.
Specifically, the rational programmer must come with \emph{modes}
that represent the alternative interpretations of blame. 
For instance, a programmer using transient may
pick the oldest element of a blame set as the next component to add types because
it corresponds to the earliest point in the program's evaluation that can
discover an impedance mismatch.  Another mode does not distinguish between the elements of the
blame set and has the rational programmer type all blamed components
at once. Because this second kind of step 
is more laborious than the first one, the analysis of the results must take 
into account the rational programmer's effort.

>>>>

The second big challenge is the disparity of gradually
typed languages. While the approaches are well understood in theory,
they come with vastly different type systems, accommodate seemingly
incompatible language features, and apply to radically different coding
styles in the untyped world.  To eliminate the effect of these differences 
on its results, the application of an evaluation method to different
gradual typing systems must use the same test programs and the same type
system. Otherwise, results for different gradual typing systems 
do not share a common baseline and, therefore, do not lend support to any
conclusions about whether the value of blame changes from one system to
another. For instance, the case study presented here builds on
Greenman and Felleisen's transient variant for functional Typed
Racket~\cite{gf-icfp-2018} to obtain a transient semantics that is
comparable with Typed Racket's default natural one. This transient
variant accommodates varied linguistic features and library calls
in a way that is faithful to the transient
semantics~\cite{vss-popl-2017}.

@@

Supporting these strategies is {\em not\/} just an issue of language
implementation but affects greatly the behavior of the rational
programmer.  Specifically, the rational programmer must come with
\emph{modes} that represent the alternative interpretations of blame.  For
instance, a mode of the rational programmer using transient may pick the
oldest element of a blame set as the next component to add types because
it corresponds to the earliest point in the program's evaluation that can
discover an impedance mismatch.  Another mode may opt to treat the blame
set as another form of a stack trace and add types to its newest element.
When both modes are equally successful in locating an impedance mismatch,
the method has needs to take into account the rational programmer's effort
to answer which mode is the most effective.

%Another mode does not distinguish between the elements of the
%blame set and has the rational programmer type all blamed components
%at once. Because this second kind of step 
%is more laborious than the first one, the analysis of the results must take 
%into account the rational programmer's effort.

### 2020-11-15 writing/mutate.tex
working on rational

  \texttt{acquire} & Board game simulation (M. Felleisen)  \\%[1em]

@@

\hline
  \texttt{take5} & Card game simulator (M.Felleisen)  \\%[1em]

@@

However,  unlike the hard-to-kill mistakes introduced by standard mutators, 
gradual type systems flag mere typos.  Thus, in order to
exercise the rational programmer in the context of gradual typing, 
we need to develop alternative mutators
that are fine-tuned to produce type-level mistakes gradual type systems 
can detect. Our inspiration is our
more-than-a-decade long experience of making type-level mistakes in Typed
Racket. Some of these mistakes often take non-trivial effort to debug and
have become the seed for our mutators, which we summarize in
figure~\ref{table:mutation-ops}.  For each one, the figure provides a
short description and an example mutation.

@@

The astute reader must have observed that mutators that produce 
type-level mistakes do not necessarily turn our benchmarks into
interesting debugging scenarios for the rational programmer.  The
benchmarks are fully typed, and type checking detects each mistake at
compile time.  Hence, we apply the mutators to versions, dubbed
\dubbed{scenarios}, of the benchmarks where some components are
stripped of their type annotation.  Without the annotations, the type
checker may not detect a mistake at compile time and, instead, the run
time checks
of the gradual type system signal an error.  Scenarios
with type-level mistakes that result in run time errors are the ideal
ground to exercise the rational programmer's ability to locate errors.

@@

The third criterion aims to capture debugging scenarios with a specific
kind of run time error: one produced by the interaction of a buggy
component with the rest of the program.  In those cases, the rational programmer
may need to examine several components to locate the source of the faulty interaction.
In technical terms, we mark these scenarios by
analyzing the stack trace of a run time error.  If the stack trace mentions at
least two distinct components, then this is accurate evidence of interaction between
components.  We require a third component because our benchmarks come
with a driver component which is present in the stack trace of all run
time errors.

>>>>

  \texttt{acquire} & Object-oriented board game implementation (M. Felleisen)  \\%[1em]

@@

\hline
  \texttt{take5} & Mixin-based card game simulator (M.Felleisen)  \\%[1em]

@@

However,  unlike the hard-to-kill mistakes introduced by standard mutators, 
gradual type systems flag mere typos.  Thus, in order to
exercise the rational programmer in the context of gradual typing, 
we need to develop alternative mutators
that are fine-tuned to produce the type-level mistakes gradual type systems 
can detect. Our inspiration is our
more-than-a-decade long experience of making type-level mistakes in Typed
Racket. Some of these mistakes often take non-trivial effort to debug and
have become the seed for our mutators, which we summarize in
figure~\ref{table:mutation-ops}.  For each one, the figure provides a
short description and an example mutation.

@@

The astute reader must have observed that mutators that produce 
type-level mistakes do not necessarily turn our benchmarks into
interesting debugging scenarios for the rational programmer.  The
benchmarks are fully typed, and type checking detects each mistake at
compile time.  Hence, we apply the mutators to versions, dubbed
\dubbed{scenarios}, of the benchmarks without 
some  of their type annotation.  Without the annotations, the type
checker may not detect a mistake at compile time and, instead, the run
time checks
of the gradual type system signal an error.  Scenarios
with type-level mistakes that result in run time errors are the ideal
ground to exercise the rational programmer's ability to locate errors.

@@

The third criterion aims to capture debugging scenarios with a specific
kind of run time error: one produced because of the interaction between the buggy
portion of the program with the rest of the program.  In those cases, the rational programmer
may need to examine several pieces of the program to locate the source of the faulty interaction.
In technical terms, we mark these scenarios by
analyzing the stack trace of a run time error.  If the stack trace mentions at
least two distinct components, then this is accurate evidence of
interaction.  We require a third component because our benchmarks come
with a driver component which is present in the stack trace of all run
time errors.

### 2020-11-15 writing/rational.tex
working on rational

Based on this idea, a metric for migratory paths is a concise way to
evaluate the effectiveness of a rational programmer. The components of
an ill-typed program, such as one of  our ill-typed mutants, 
create a \emph{debugging scenario space}, a lattice like
the one introduced by ~\citet{tfgnvf-popl-2016} for performance
evaluations. As we discuss briefly at the end of section ~\ref{sec:mutate}, 
this space consists of variants of the program, dubbed
\emph{scenarios}, that differ in their type annotations. In the
setting of migratory typing,\footnote{The method can be adapted to
so-called true gradual typing systems~\cite{svcb-snapl-2015} in a
straightforward manner.} each scenario differs from another in which
components are typed. Formally, a program $\system$ is a set of
components $\set{\component}$, and a scenario $\conf$ is a subset of
$\set{\component}$ . These scenarios of $\system$ are ordered by the
subset relation and thus form a lattice $\lattice{\system}$ with
$2^{\size{\system}}$ elements.  The bottom scenario of
$\lattice{\system}$ is always $\emptyset$ and the top one is $\system$
itself. In between are all the mix-typed scenarios.

With these definitions, the actions of the rational programmer map to
an ascending chain---dubbed a \emph{trail}---in $\lattice{\system}$.
The starting scenario $\conf_0$ of a trail---also referred to as the
\emph{root} (of the trail)---is the scenario that the programmer
attempts to debug initially. The easiest scenarios are those that the
type checker rejects outright, because a type error pinpoints an
impedance mismatch between the broken code of one component and its
typed interface to others. At this point, the programmer has
identified the broken component. If the scenario type-checks, the
programmers runs the program until it raises an exception.  The
rational programmer uses the blame assignment\footnote{Blame can
result either directly from a run time type check or, in the absence
of blame, from an exception that we track back to the component that
causes it with the help of the stack trace.} to decide which
component(s) to equip with types. This choice constructs scenario
$\conf_{i+1}$ from $\conf_{i}$ and implies the creation of an
ascending chain.  Intuitively, with each step along the chain, the
rational programmer reduces the current scenario to an ``easier'' one
to debug.

@@

The next three subsections specialize the trail-scenario idea to the
various semantics of gradual typing. Once this machinery is in place,
we use the fourth subsection to state the precise experimental
research questions that correspond to the high-level question on the
first page and to explain our experimental process. The final
subsection extends the development with a discussion of how we can
measure the rational programmer's effort along the trail.

@@

When the buggy untyped component of a program is replaced by the typed counterpart,
the type checker fails if this component causes the impedance mismatch. Hence, a
trail that ensd at an ill-typed scenarios successfully pinpoints the location of
bugs. 
\begin{quote}
\it A natural blame trail $\conf_0,...\conf_n$ in a lattice $\lattice{\system}$ is
\emph{successful} iff its last scenario $\conf_n$ does not type check.  A natural
blame trail $\conf_0,..,\conf_n$ in a lattice $\lattice{\system}$ is \emph{failing}
iff $\conf_n$ type checks and the trail cannot be extended further.
\end{quote}
That is failing natural blame trails are those that have not reached a scenario that
reveals the bug at compile time and their last scenario does not produce blame. Thus
the rational programmer receives no hints from the gradual type system on how to
continue the search for the bug.

Equipped with these preliminary definitions, we can precisely define what it means
for blame to add value to a gradual type system in a specific setting. 
\begin{quote}
\it
Given a program $\system$, natural blame is\\
(1) \emph{always useful} iff $\lattice{\system}$ contains no failing natural blame trails.\\
(2) \emph{sometimes useful} iff $\lattice{\system}$ contains successful natural blame trails.\\
(3) \emph{never useful} iff $\lattice{\system}$ contains no successful natural blame
trails.\\
\end{quote}

While these definitions describe how a rational programmer might gauge whether it
pays off to heed blame assignments during debugging, they do not answer whether
blame is a critical piece of the rational programmer's process.  For instance,
typing a random component might be as useful as typing the blamed one.

To account for this situation, we devise a new mode of the natural rational
programmer that follows a random migration process:
\begin{quote}
\it A natural random trail is a sequence of scenarios $\conf_0,...\conf_n$ of a
program $\system$ such that for all $0 \leq i \leq n - 1$, $\conf_i \subset
\conf_{i+1}$ and $\conf_{i+1} \setminus \conf_i = \{\random{\system, \conf_i}\}$
where $\random{\system, \conf}$ denotes a randomly selected component of $\system$
that is not in $\conf_i$.
\end{quote}
Similar to natural blame we determine how successful the random
migration is based on the success of natural random trails: 
\begin{quote}
\it
Given a program $\system$, natural random blame is\\
(1) \emph{always useful} iff $\lattice{\system}$ contains no failing natural random blame trails.\\
(2) \emph{sometimes useful} iff $\lattice{\system}$ contains successful natural random blame trails.\\
(3) \emph{never useful} iff $\lattice{\system}$ contains no successful natural
random blame trails.\\
\end{quote}

Putting the pieces together, we can now define that blame assignments adds value in
the context of a natural semantics if blame-based migration is better than random
migration.
\begin{quote}
\it Natural blame adds value to natural for a program $\system$ over pure chance iff
  natural blame is more successful than natural random. And, natural blame is more
  successful than natural random for $\system$ if either natural blame is always
  successful for $\system$ when natural random is sometimes or never successful, or
  natural blame is sometimes successful for $\system$ when natural random is never
  successful.
\end{quote}

 Since gradually typed languages with erasure semantics do not come with
 blame assignment, a rational programmer can only hope that the underlying
 safety checks and their exceptions are helpful.  Thus, adapting the
 definitions related to exception and random trails from natural to
 erasure, we characterize when the effect of erasure exceptions in debugging a
 program $\system$ goes beyond that of pure chance.

\begin{quote}
\it
  Erasure exceptions add value to erasure for a program $\system$ over pure chance
  iff erasure exception is more successful than erasure random.
\end{quote}
\noindent

@@

This ambiguity in transient blame raises the question of how the rational programmer
should react when the language produces a blame set. Our answer is that the rational
programmer has at least two reasonable options.  The first option is to type all
blamed components at once. The second one is to select the component that is added
to the blame set first and assign types to only that one---after all the types of
this first component should be able to detect an impedance mismatch earlier in the
evaluation of a program than the later ones.

These two modes of rationalizing give rise to two different notions of trail.
\begin{quote}
\it A \emph{ transient-all blame trail} is a sequence of scenarios
$\conf_0,...\conf_n$ of a program $\system$ where for all $0 \leq i \leq n - 1$,
$\conf_i \subset \conf_{i+1}$ and ~$\conf_{i+1} \setminus \conf_i = \mblame{\system,
\conf_i}$ where $\mblame{\system, \conf}$ denotes the set of components that $\conf$
blames under transient semantics.
\end{quote}
\noindent
and
\begin{quote}
\it A \emph{transient-firstborn blame trail} is a sequence of scenarios
$\conf_0,...\conf_n$ of a program $\system$ where for all $0 \leq i \leq n - 1$,
$\conf_i \subset \conf_{i+1}$ and $\conf_{i+1} \setminus \conf_i =
\first{\mblame{\system, \conf_i}}$ where $\first{\mblame{\system, \conf}}$ is the
first component transient adds to the blame set for $\conf$.
\end{quote}
The definitions of \emph{transient erasure trails} and \emph{transient random
trails} are analogous to those for natural erasure trails and natural random trails,
respectively.

We also use these four forms of blame trails to define when transient-all and
transient-firstborn add value to transient over exceptions and pure chance.
\begin{quote}
\it 
Transient all blame adds value to transient for a program $\system$ over
  pure chance and transient exceptions iff transient all blame 
  is more successful than both transient random and transient exception.

Transient firstborn blame adds value to transient for a program $\system$ over
  pure chance and transient exceptions iff transient all blame 
  is more successful than both transient random and transient exception.

\end{quote}

@@

Trails and their properties give us the tools for a rigorous examination
of blame for natural, transient and erasure Typed Racket in the setting of the 
mutants from section~\ref{sec:mutate}. In particular, our experimental
process collects data to answer four questions for each mutant:
\begin{itemize}
\item[$Q_1$] Is blame valuable for natural?

\item[$Q_2$] Is all blame valuable for transient?

\item[$Q_3$] Is firstborn blame valuable for transient?

\item[$Q_*$] Are plain exceptions valuable for
  erasure? For the other two systems? 
\end{itemize}

Table~\ref{fig:experiment-outline} summarizes how each question relates to
different kinds of trails/modes of the rational programmer. For example experimental
question $Q_1$ asks whether blame is valuable for natural and our experiment
uses the natural blame, exception and random trails to answer it.

\begin{figure}[ht]
\center
{\begin{tabular}{l|c|c|c}
%% ---------------------------------------------------------------------------------------------------------------
                        & {\bf Natural}        & {\bf Transient}          & {\bf Erasure} \\ \hline 
%% ---------------------------------------------------------------------------------------------------------------
{\bf Blame}             &       $Q_1$          &                          &               \\
{\bf All blame}         &                      &     $Q_2$                &               \\
{\bf Firstborn blame}   &                      &     $Q_3$                &               \\
{\bf Exception}         &       $Q_1$/$Q_*$    &     $Q_2$/$Q_3$/$Q_*$    &      $Q_*$    \\
{\bf Random}            &       $Q_1$/$Q_*$    &     $Q_2$/$Q_3$/$Q_*$    &      $Q_*$    \\
\end{tabular}}
  \caption{ Experimental questions and rational programmer modes.}
  \label{fig:experiment-outline}
\end{figure}

@@

In detail we answer $Q_1$ for an ill-typed mutant from
section~\ref{sec:mutate} by creating trails for each natural blame,
natural exception and natural random trail in the scenario lattice of the mutant. A
log reports on success and failure endings.  The first step is to construct the
mutant's scenario lattice and select an interesting debugging scenario
as we discuss in section~\red{sec:mutate} to be the root of a trail. Our test
bed extends these trails according to the natural-blame programmer.  If no scenarios
can be added to the trail---a trail may the top of the finite lattice---we check if
the last scenario of the trail type checks or not. If it does, this natural-blame
trail is successful, otherwise it is failing. We repeat the same process for the
roots two more times; once following the natural-exception programmer and once the
natural-random programmer. Finally, based on the collected data, we calculate the
success/failure results for the trails to determine whether natural blame is more
successful than natural exceptions and pure chance and we obtain the answer for
$Q_1$.  The process is analogous for the other questions employing the respective
modes of the rational programmer. Figure~\ref{fig:process} summarizes the
experimental process for one mode of the rational programmer and connects
it with the mutations from section~\ref{sec:mutation}.

@@

\subsection{Programmer Effort}

@@

Programmer effort can help shed some light to the relative effectiveness of the
three gradual typing systems. For example, consider a debugging scenario from  the
lattice of one of our mutants.  If the
scenario results in blame both under, for instance, the natural and transient semantics,
then the natural blame rational programmer and the two transient blame
rational programmers (all and firstborn) can compete to see which one
reaches an ill-typed scenarion with less effort. In
general, the effort of trails of different modes that have the same
root is a point of comparison for the effectiveness of these
modes even if the modes correspond to different gradual typing systems.
Unfortunately, statistically speaking, such comparisons between two modes do not generalize 
beyond the successful trails with common roots in our experiments.
Still they allow us to quantify the difference between the modes of the rational
programmers and thus the different blame assignment strategies  for the specific
debugging scenarios we analyze in our experiment.

>>>>

Based on this idea, migratory paths offer a concise setting for
evaluating the effectiveness of a rational programmer. If we group the
scenarios from section ~\ref{sec:mutate} by mutation, then we obtain a set 
of programs that all contain the same mistake but differ in their type
annotations.  In the setting of migratory typing,\footnote{The method can be adapted to
so-called true gradual typing systems~\cite{svcb-snapl-2015} in a
straightforward manner.} each scenario differs from another in which
components are typed. Formally, a program $\system$ is a set of
components $\set{\component}$, and a scenario $\conf$ is the typed subset of
$\set{\component}$ . Hence, the scenarios of $\system$ are ordered by the
subset relation and, naturally,  form a lattice $\lattice{\system}$ with
$2^{\size{\system}}$ elements, like
the one introduced by ~\citet{tfgnvf-popl-2016} for performance
evaluations. The bottom scenario of
$\lattice{\system}$ is always $\emptyset$ and the top one is $\system$
itself. In between are all the mix-typed scenarios.

With these definitions, the actions of the rational programmer map to an
ascending chain---dubbed a \emph{trail}---in $\lattice{\system}$.  The
starting scenario $\conf_0$ of a trail---also referred to as the
\emph{root} (of the trail)---is the initial debugging scenario.  Herein we
consider only roots that are interesting debugging scenarios as described
in~\ref{sec:mutate} --- we return to this point further on.  Intuitively,
starting from such a root, the rational programmer  steps along a trail,
and at each step, uses reduces the current scenario to an ``easier'' next
scenario to debug.  The easiest scenarios are those that the type checker
rejects outright, because a type error pinpoints an impedance mismatch
between the broken code of one component and its typed interface to
others. At this point, the programmer has identified the broken component.
If the scenario type-checks, the programmers runs the program until it
raises an exception.  The rational programmer uses the blame
assignment\footnote{Blame can result either directly from a run time type
check or, in the absence of blame, from an exception that we track back to
the component that causes it with the help of the stack trace.} to decide
which component(s) to equip with types. This choice constructs scenario
$\conf_{i+1}$ from $\conf_{i}$ and implies the creation of trail.

@@

The next three subsections specialize the trail-scenario idea to the
various semantics of gradual typing. Once this machinery is in place,
we use the fourth subsection to state the precise experimental
research questions that correspond to the high-level question on the
first page and to explain our experimental process. The final
subsection extends the development with a discussion of how we can
compare the effectiveness of the rational programmer across semantics.

@@

When the buggy untyped component of a program is replaced by the typed counterpart,
the type checker fails if this component causes the impedance mismatch. Hence, a
trail that ends at an ill-typed scenarios successfully pinpoints the location of
bugs. 
\begin{quote}
\it A natural blame trail $\conf_0,...\conf_n$ in a lattice $\lattice{\system}$ is
\emph{successful} iff its last scenario $\conf_n$ does not type check.  A natural
blame trail $\conf_0,..,\conf_n$ in a lattice $\lattice{\system}$ is \emph{failing}
iff $\conf_n$ type checks and the trail cannot be extended further.
\end{quote}
That is failing natural blame trails are those that have not reached a scenario that
reveals the bug at compile time and their last scenario does not produce blame. Thus
the rational programmer receives no hints from the gradual type system on how to
continue the search for the bug.

Equipped with these preliminary definitions, we can precisely define what it means
for blame to add value to a gradual type system in a specific setting. 
\begin{quote}
\it
Given a program $\system$, the natural blame mode is\\
(1) \emph{always useful} iff $\lattice{\system}$ contains no failing natural blame trails.\\
(2) \emph{sometimes useful} iff $\lattice{\system}$ contains successful natural blame trails.\\
(3) \emph{never useful} iff $\lattice{\system}$ contains no successful natural blame
trails.\\
\end{quote}

While these definitions describe how a rational programmer might gauge whether it
pays off to heed blame assignments during debugging, they do not answer whether
blame is a critical piece of the rational programmer's process.  For instance,
typing the top of stack trace of a failed run time type check might be as useful as typing the blamed one.

To account for this situation, we devise a new mode of the natural rational
programmer that follows a migration process based on exceptions:
\begin{quote}
\it A natural exception trail is a sequence of scenarios $\conf_0,...\conf_n$ of a
program $\system$ such that for all $0 \leq i \leq n - 1$, $\conf_i \subset
\conf_{i+1}$ and $\conf_{i+1} \setminus \conf_i = \{\random{\system, \conf_i}\}$
where $\exception{\system, \conf}$ denotes the location of the runtime
  error of $\conf$.
\end{quote}
Similar to natural blame we determine how successful the exception
migration is based on the success of natural exception trails: 
\begin{quote}
\it
Given a program $\system$, the natural exception mode is\\
(1) \emph{always useful} iff $\lattice{\system}$ contains no failing
  natural exception trails.\\
(2) \emph{sometimes useful} iff $\lattice{\system}$ contains successful
  natural exception trails.\\
(3) \emph{never useful} iff $\lattice{\system}$ contains no successful natural
exception trails.\\
\end{quote}

Putting the pieces together, we can now define that blame assignments adds value in
the context of a natural semantics if blame mode  is better than the
exception.
\begin{quote}
\it Natural blame adds value to natural for a program $\system$ over
  exceptions iff the natural blame mode is more successful than the
  natural exception mode. And, the natural blame mode is more successful
  than the natural exception mode for $\system$ if either natural blame is
  always successful for $\system$ when natural exception is sometimes or
  never successful, or natural blame is sometimes successful for $\system$
  when natural exception is never successful.
\end{quote}

@@

This ambiguity in transient blame raises the question of how the rational programmer
should react when the language produces a blame set. Our answer is that the rational
programmer has at least two reasonable options. The first one is to select the component that is added
to the blame set first and assign types to only that one---after all the types of
this first component should be able to detect an impedance mismatch earlier in the
evaluation of a program than the later ones. The second option is to
select the component that is added to the blame set last, effectively
interpreting the blame set as an enhanced stack trace.

These two modes of rationalizing give rise to two different notions of trail.
\begin{quote}
\it A \emph{transient-first blame trail} is a sequence of scenarios
$\conf_0,...\conf_n$ of $\system$ where for all $0 \leq i \leq n - 1$,
$\conf_i \subset \conf_{i+1}$ and $\conf_{i+1} \setminus \conf_i =
\first{\mblame{\system, \conf_i}}$ where $\first{\mblame{\system, \conf}}$ is the
first component transient adds to the blame set for $\conf$.
\end{quote}

and

\begin{quote}
\it A \emph{transient-last blame trail} is a sequence of scenarios
$\conf_0,...\conf_n$ of $\system$ where for all $0 \leq i \leq n - 1$,
$\conf_i \subset \conf_{i+1}$ and $\conf_{i+1} \setminus \conf_i =
  \first{\mblame{\system, \conf_i}}$ where $\last{\mblame{\system, \conf}}$ is the
last component transient adds to the blame set for $\conf$.
\end{quote}

The definition of \emph{transient exception trails} 
is analogous to that for natural and, same as for natural, we use it to  
specify when the transient-last and transient-first blame modes add
value to transient for a program.

 Since gradually typed languages with erasure semantics do not come with
 blame assignment, a rational programmer can only hope that the underlying
 safety checks and their exceptions are helpful.  Thus, the erasure 
 rational programmer has a single mode; the erasure exception mode.

@@

Trails and their properties give us the tools for a rigorous examination
of blame for natural, transient and erasure Typed Racket in the setting of the 
mutants from section~\ref{sec:mutate}. In particular, our experimental
process collects data to answer three questions for each mutant:
\begin{itemize}
\item[$Q_1$] Is blame valuable for natural?

\item[$Q_2$] Is first blame valuable for transient?

\item[$Q_3$] Is last blame valuable for transient?

\end{itemize}

Table~\ref{fig:experiment-outline} summarizes how each question relates to
different kinds of trails/modes of the rational programmer. For example experimental
question $Q_1$ asks whether blame is valuable for natural and our experiment
uses the natural blame and exception trails to answer it.

\begin{figure}[ht]
\center
{\begin{tabular}{l|c|c}
%% ---------------------------------------------------------------------------------------------------------------
                        & {\bf Natural}        & {\bf Transient}         \\ \hline 
%% ---------------------------------------------------------------------------------------------------------------
{\bf Blame}             &       $Q_1$          &                           \\
{\bf First blame}         &                      &     $Q_2$                \\
{\bf Last blame}   &                      &     $      Q_3$                  \\
{\bf Exceptions}         &       $Q_1$    &     $Q_2$/$Q_3$    \\
\end{tabular}}
  \caption{ Experimental questions and rational programmer modes.}
  \label{fig:experiment-outline}
\end{figure}

@@

In detail we answer $Q_1$ for an ill-typed mutant from
section~\ref{sec:mutate} by analyzing the natural blame and
natural exception trails in the scenario lattice of the mutant. A
log reports on success and failure endings.  The first step is to construct the
mutant's scenario lattice and select an interesting debugging scenario
as  the root of a trail.
Our test
bed extends these trails according to the natural-blame programmer.  If no scenarios
can be added to the trail---a trail may the top of the finite lattice---we check if
the last scenario of the trail type checks or not. If it does, this natural-blame
trail is successful, otherwise it is failing. We repeat the same process for the
same roots once more following the natural-exception programmer
. Finally, based on the collected data, we calculate the
success/failure results for the trails to determine whether natural blame is more
successful than natural exceptions  and we obtain the answer for
$Q_1$.  The process is analogous for the other questions employing the respective
modes of the rational programmer. Figure~\ref{fig:process} summarizes the
experimental process for one mode of the rational programmer and connects
it with the mutations from section~\ref{sec:mutate}.

@@

\subsection{Comparing Modes through Programmer Effort}

@@

Programmer effort can help shed some light to the relative effectiveness of the
three gradual typing systems. For example, consider a debugging scenario from  the
lattice of one of our mutants.  If the
scenario results in blame both under, for instance, the natural and transient semantics,
then the natural blame rational programmer and the two transient blame
rational programmers (all and firstborn) can compete to see which one
reaches an ill-typed scenario with less effort. In
general, the effort of trails of different modes that have the same
root is a point of comparison for the effectiveness of these
modes even if the modes correspond to different gradual typing systems.
Since in our experiment all roots are interesting debugging scenarios and 
those are independent of mode or semantics, effort of trails is sufficient 
to compare the relative value of one mode versus another.
